NumPy---Array Computation for Python

### 2019-12-20 body.tex
Remove cruft
Revise intro

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
It underpins almost every library that does numerical
computation, including SciPy.  NumPy consists of the \code{ndarray}-object
along with utility functions to operate on it.  Because of
its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy
array is the {\it de facto} exchange format for array data in
Python.
NumPy, while having been wildly successful as a library and array exchange
project, has now also become a standard language for expressing array and
tensor computations.  
It has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous, and is
implemented by, for example, popular tensor computation libraries.

The arrival of a stable version of NumPy (1.0) in October 2006 unified
the scientific Python community, and heralded the era of popular
scientific computing in Python that has led to formidable successes
such as [cite some impacts here].

[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
  IDL is now below Python, Pangeo <-> xarray, Earthcube,
  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

@@

NumPy was, at its inception, the unifier of the Python array
processing world.  Before NumPy, two packages—Numeric and
numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
small arrays.  In contrast, to handle large astronomical images,
NumArray was developed at the Space Telescope Science Institute (STScI).
In other words, these array
objects had different strengths and weaknesses, and therefore tended
to divide the already small user community.  Furthermore, the divide
posed a challenge for SciPy, which built on Numeric and provided
fundamental scientific routines \cite{virtanen2019scipy}.
In response, the maintainer of Numeric, Travis Oliphant, started a
major rewrite that aimed to be a ``best of both worlds'' unification
of the two projects.  This became known as NumPy.

While, by 2006, NumPy was already eminently usable, it lacked good
documentation. In order to use NumPy and SciPy in his classes, Joe Harrington
from the University of South Florida sponsored several rounds of the
NumPy Documentation Marathon \cite{harrington2008scipy,vanderwalt2008scipy}.
During these intensive
documentation improvement sessions, we developed a Wikipedia-like
system for community members to contribute so-called
``docstrings''—text describing each NumPy function. A community
documentation standard was published so that these pieces of
documentation would all provide similar information, and conform to a
basic standard of quality. Experienced community members reviewed and
integrated contributed docstrings, and soon NumPy had thorough
documentation coverage.

To coordinate work as a global community, to ensure accuracy of computed
results, and to allow modification without breakage, the NumPy community
pioneered best software engineering practices in the scientific Python
community  \cite{millman2014developing}.
For example, NumPy early on adopted distributed revision control, unit testing,
continuous integration, code review, and standardized executable documentation.

However, after his initial effort to build NumPy, Travis Oliphant
stepped back from development to found a company and several projects,
including Numba---an accelerated Python compiler that could vastly
accelerate NumPy expressions.  By 2009, the project had entered a
period of uncoordinated, opportunistic development, and was
increasingly weighed down by technical debt that had accrued since the
days of Numeric.

@@

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the
package had rapid adoption and became central to scientific computing
without external investment.

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development

In 2017, the Berkeley Institute for Data Science
(BIDS) received grants totaling 1.3M USD from the Gordon \& Betty
Moore and the Alfred P. Sloan foundations to support the development
of NumPy for better data science.
Specifically, the aim was to address technical debt and set in place
standards and architecture to encourage more sustainable development.
Stéfan van der Walt, a senior research data scientist at BIDS and a
member of the NumPy steering council, is the PI and manages four
programmers working on the project.

% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/

In 2019, NumFOCUS received a grant for \$195,000 in 2019 for the
development of NumPy and OpenBLAS (on which NumPy depends for
accelerated linear algebra) from the Chan Zuckerberg Initiative.
This grant, building on the technical work supported by the previous
grant, will focus on better serving NumPy's large number of beginner
to intermediate level users and on growing the community of NumPy
contributors.
Ralf Gommers, the Director of Quansight Labs and a member of the NumPy
steering council, is the PI.

The project receives around 10K USD per year from Tidelift, which is
used to fund documentation and website developers.

@@

In 2007, the NumPy release manager, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently until 2017 when, to
facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
Since then there have been 19 proposed NEPS---6 have been implemented,
4 have been accepted and are being implemented, 4 are under
consideration, 3 have been deferred or superseded, and 2 have been rejected
or withdrawn.

@@

% \subsection*{GSoD}

@@

% use scipy for linear algebra etc...

%A suite of functions for computing the \emph{fast Fourier transform (FFT)}
%and its inverse is provided.

%NumPy's linear algebra library includes functions for: solving linear
%systems of equations; computing various functions of a matrix, including
%the determinant, the norm, the inverse, and the pseudo-inverse;
%computing the Cholesky, eigenvalue, and singular value decompositions of a matrix;
%and more.
%

%For historical reasons, we provide a \code{matrix} class the implements 2-d
%arrays with specialized behavior.
%But we strongly recommend you use a 2-d array and use the infix \code{@} operator
%for matrix multiplication.
%
%There is also a \code{matrix} class the implements 2-d arrays with
%specialized behavior.  We mention it here for completeness, but the
%use of \code{matrix} is discouraged.  The quirks of the \code{matrix}
%class have proven to be troublesome, and with the addition of the
%infix \code{@} operator to Python for matrix multiplication, there was no
%longer a need for a separate \code{matrix} class, since one can
%use a 2-d array instead.

@@

% \section*{Common infrastructure}

%\subsection*{Wheels build system}
%
%A Python \emph{wheel}\cite{PEP427} is a standard file format for
%distributing Python libraries.  In addition to Python code, a
%wheel may include compiled C extensions and other binary data.
%This is important, because many libraries, including NumPy,
%require a C compiler and other build tools to build the software
%from the source code, making it difficult for many users to install
%the software on their own.  The introduction of wheels to the Python
%packaging system has made it much easier for users to install
%precompiled libraries.
%
%A GitHub repository containing scripts to build NumPy wheels has
%been configured so that a simple commit to the repository triggers
%an automated build system that creates NumPy wheels for several
%computer platforms, including Windows, Mac OSX and Linux.  The wheels
%are uploaded to a public server and made available for anyone to use.
%This system makes it easy for users to install precompiled versions
%of NumPy on these platforms.
%
%The technology that is used to build the wheels evolves continually.
%At the time this paper is being written, a key component is the
%\code{multibuild} suite of tools developed by Matthew Brett and
%other developers\cite{multibuild}.  Currently, scripts using
%\code{multibuild} are written for the continuous integration
%platforms Travis-CI (for Linux and Mac OSX) and Appveyor
%(for Windows).

@@

% started at as focused small matrices, class examples

>>>>

NumPy is the fundamental array computation library for the Python
ecosystem \cite{dubois2007guest,oliphant2007python,millman2011python,perez2011python}.
Since its first stable release (1.0) in October 2006 it has unified
the scientific Python community and now underpins almost every library
that does numerical computation, including SciPy\cite{virtanen2019scipy},
pandas\cite{mckinney-proc-scipy-2010}, scikit-learn\cite{pedregosa2011scikit},
and scikit-image\cite{vanderwalt2014scikit}.
Not surprisingly, many important scientific research projects and
results rely on NumPy.
%[LIGO, EHT, Higgs-Boson pipeline at CERN—double check this carefully,
%  may be root, ice cube experiment, https://arxiv.org/abs/1507.03989
%  IDL is now below Python, Pangeo <-> xarray, Earthcube,
%  https://ai.jpl.nasa.gov/public/documents/papers/hackett\_spaceops2018\_block.pdf,
%  Kepler, Mars rovers, SKA in South Africa, yt, astropy, seismo-live,
%  icesat2 \& icepyx, The Materials Project at LBNL https://github.com/materialsproject]

NumPy consists of the \code{ndarray}-object along with utility functions to
operate on it.
Because of its inherent simplicity—being a pointer to memory with some
associated meta-data about shape, data-type, and so forth—the NumPy array is
the {\it de facto} exchange format for array data in Python.
NumPy, while having been wildly successful as a library and array exchange
project, has now also become a standard language for expressing array and
tensor computations.  
It has such widespread adoption that not only the array object but also its
{\it Application Programming Interface} (API) has become ubiquitous, and is
implemented by, for example, popular tensor computation libraries.

@@

Before NumPy, two different Python array packages—Numeric and
numarray—existed.  The first was developed by Jim Fulton, Jim Hugunin, Paul Dubois, Konrad
Hinsen, Travis Oliphant, and others, and was aimed at fast manipulation of many
small arrays.  In contrast, to handle large astronomical images,
NumArray was developed at the Space Telescope Science Institute (STScI).
These array objects had different strengths and weaknesses, and therefore
divided the already small community.  Furthermore, the divide
posed a challenge for SciPy, which was built on Numeric and provided
fundamental scientific routines.
In response, the maintainer of Numeric, Travis Oliphant, started a
major rewrite that aimed to be a ``best of both worlds'' unification
of the two projects.  This became known as NumPy.

This spurred renewed development in both SciPy and NumPy and 
heralded the current era of wide-spread use of Python for scientific computing.
To coordinate work as a global community, to ensure accuracy of computed
results, and to allow modification without breakage, the NumPy (and SciPy) community
pioneered best software engineering practices in the scientific Python
community  \cite{millman2014developing}.
For example, NumPy early on adopted distributed revision control, unit testing,
continuous integration, code review, and standardized executable documentation.
It also developed a standard for ``docstrings''—text describing each NumPy function.
This quickly became the standard for the larger scientific Python community.

As the scientific Python community grew and new challenges arose, Travis Oliphant
eventually stepped back from leading NumPy to found a company and start
new projects including Numba---an accelerated Python compiler that could vastly
accelerate NumPy expressions.
Development of NumPy continued by a core group that informally operated by consensus.

%By 2009, the project had entered a
%period of uncoordinated, opportunistic development, and was
%increasingly weighed down by technical debt that had accrued since the
%days of Numeric.

@@

For many years, NumPy had no dedicated funding, development
time was mostly contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the
package had rapid adoption and became central to scientific computing
without external investment.

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development
In 2017, NumPy received its first large grants totaling 1.3M USD from the
Gordon \& Betty Moore and the Alfred P. Sloan foundations.
Stéfan van der Walt is the PI and manages four programmers working on the project.
These two grants focus on addressing the technical debt accrued over the years and
on setting in place standards and architecture to encourage more sustainable development.

% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/
NumPy received another grant for \$195,000 from the Chan Zuckerberg
Initiative at the end of 2019.
Ralf Gommers is the PI.
This grant, building on the technical work supported by the previous
grant, will focus on better serving NumPy's large number of beginner
to intermediate level users and on growing the community of NumPy
contributors.
It also will provide some support for OpenBLAS (on which NumPy depends for
accelerated linear algebra).

The project also receives around 10K USD per year from Tidelift, which is
used to fund documentation and website developers.

@@

As the NumPy release manager in 2007, Jarrod Millman, adopted the notion
of enhancement proposals---used infrequently and informally until 2017 when,
to facilitate work expected from funded development, he formalized the
NumPy Enhancement Proposal (NEP) process.  NEPs are modeled after
Python Enhancement Proposals (PEPs) for ``proposing major new
features, for collecting community input on an issue, and for
documenting the design decisions that have gone into
Python''\footnote{\url{https://numpy.org/neps/nep-0000.html}}.
Since then there have been 19 proposed NEPS---6 have been implemented,
4 have been accepted and are being implemented, 4 are under
consideration, 3 have been deferred or superseded, and 2 have been rejected
or withdrawn.

@@

@@

@@

@@
